Partial Least Squares Regression with Multiple Functional and Scalar Predictors

### 2025-06-29 manuscript/main.tex
proposition 1 revised

For the computation of the estimated PLS direction  $\hat{\boldsymbol{\xi}}$, 
we obtain the basis coefficients of its functional part, denoted by $\mathbf{d}^{(1)}, \ldots, \mathbf{d}^{(K)} \in \mathbb{R}^{M}$,
and the scalar part  
$\mathbf{d} \in \mathbb{R}^p$,
by solving linear systems
\begin{equation*}
 \boldsymbol{\Phi}^{(1)}
\mathbf{d}^{(1)} 
=
(\widetilde{\mathbf{C}}^{(1)} \boldsymbol{\Phi}^{(1)})^\top
\mathbf{y},
\ldots
,
\boldsymbol{\Phi}^{(K)}
\mathbf{d}^{(K)} 
=
(\widetilde{\mathbf{C}}^{(K)} \boldsymbol{\Phi}^{(K)})^\top
\mathbf{y},
~\text{and}~
\mathbf{d} := \mathbf{Z}^\top \mathbf{y} .
\end{equation*}
We then normalize these coefficients and scalar parts by the hybrid Hilbert norm (Definition
\ref{def:hilbert_space}),
which can be computed as
\begin{equation*}
\|	\hat{\boldsymbol{\xi}} \|_{\mathcal{H}} = 
\bigl(
	\mathbf{d}^\top \mathbf{d} + 
	\sum_{k=1}^K (\mathbf{d}^{(k)})^\top  \boldsymbol{\Phi}^{(K)}  \mathbf{d}^{(k)} 
	\bigr)^{1/2}.
\end{equation*}
Therefore, the estimated PLS direction is the tuple $\hat{\boldsymbol{\xi}} \in \mathcal{H}$, defined as
\begin{equation}
	\hat{\boldsymbol{\xi}} :=
	\frac{1}{
	\bigl(
	\mathbf{d}^\top \mathbf{d} + 
	\sum_{k=1}^K (\mathbf{d}^{(k)})^\top  \boldsymbol{\Phi}^{(K)}  \mathbf{d}^{(k)} 
	\bigr)^{1/2}
	}
	\biggl(
	\sum_{m=1}^M d^{(1)}_m \phi^{(1)}_m,
	\ldots,
	\sum_{m=1}^M d^{(K)}_m \phi^{(K)}_m,
	\mathbf{d}
	\biggr).
\end{equation}
Finally, the estimated $l$th PLS scores for the $n$ observations,
denoted $\hat{\boldsymbol{\rho}} \in \mathbb{R}^n$
 is computed simultaneously as the hybrid inner product (Definition \ref{def:n_one_inner_product}):
	\begin{equation*}
	\langle \widetilde{\mathbf{W}}, \hat{\boldsymbol{\xi}} \rangle_{\mathcal{H}}
	=
	\frac{1}{\|	\hat{\boldsymbol{\xi}} \|_{\mathcal{H}}}
	\biggl(
	\mathbf{Z} \mathbf{d}
	+
	\sum_{k=1}^K \widetilde{\mathbf{C}}^{(k)} \boldsymbol{\Phi}^{(k)} \mathbf{d}^{(k)}
	\biggr)
	. 
\end{equation*}

%	consider the basis function expansions of the functional predictors $X_i^{(k)}(t_k) = \sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t_k)$, $i=1, \ldots, n$, and those of the functional parts of the PLS component $\psi^{(k)}(t_k) = \sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t_k)$, $k=1, \ldots, K$. Let 
	\begin{proposition} \label{Proposition: generalized rayleigh quotient}
		Let $\operatorname{span}(B^\ast[\mathbf{t}])$ 
		be a linear subspace of $\mathcal{H}$ spanned by the basis functions.
		  For any unit-length direction  $\boldsymbol{\xi} \in \operatorname{span}(B^\ast[\mathbf{t}])$,
expressed as $
\boldsymbol{\xi} 
= 
\boldsymbol{\alpha}
B^\ast [\mathbf{t}] $ with $\boldsymbol{\alpha} \in \mathbb{R}^{MK + p}$,
		Let us define
		the empirical and basis-approximated sample covariance as
\begin{equation*}
	\widehat{\textnormal{Cov}} (\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y )
	:=
	\frac{1}{n} \sum \limits_{i=1}^n \langle \widetilde{W}_i, \boldsymbol{\xi} \rangle_\mathcal{H} y_i
	=
		\frac{1}{n} \sum \limits_{i=1}^n 
	(	\tilde{\mathbf{C}}^\ast_i \mathbf{J}^\ast \boldsymbol{\alpha} ) y_i.
\end{equation*}
Then the coefficients computed in \eqref{compute:pls_direction_coeff}  $\widehat{\textnormal{Cov}} (\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y )$
over all unit vectors in $\operatorname{span}(B^\ast[\mathbf{t}])$.
In a more formal way:
	\begin{equation}
	\begin{bmatrix} \mathbf{a} \\ \mathbf{\theta} \end{bmatrix}
	B^*[\mathbf{t}] \in \arg \max_{
 \|\boldsymbol{\xi}  \| = 1, 
	 \boldsymbol{\xi} \in \mathrm{span}(B^*[\mathbf{t}])
	}
		\widehat{\textnormal{Cov}} (\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y ).
\end{equation}

>>>>

For the computation of the estimated PLS direction  $\hat{\boldsymbol{\xi}}$, 
we obtain the basis coefficients of its functional part, denoted by $\mathbf{d}^{(1)}, \ldots, \mathbf{d}^{(K)} \in \mathbb{R}^{M}$,
and the scalar part  
$\mathbf{d} \in \mathbb{R}^p$,
by solving linear systems
\begin{equation*}
 \boldsymbol{\Phi}^{(1)}
\mathbf{d}^{(1)} 
=
(\widetilde{\mathbf{C}}^{(1)} \boldsymbol{\Phi}^{(1)})^\top
\mathbf{y},
\ldots
,
\boldsymbol{\Phi}^{(K)}
\mathbf{d}^{(K)} 
=
(\widetilde{\mathbf{C}}^{(K)} \boldsymbol{\Phi}^{(K)})^\top
\mathbf{y},
~\text{and}~
\mathbf{d} := \mathbf{Z}^\top \mathbf{y} .
\end{equation*}
We then normalize these coefficients and scalar parts by the hybrid Hilbert norm (Definition
\ref{def:hilbert_space}),
which can be computed as
\begin{equation*}
\|	\hat{\boldsymbol{\xi}} \|_{\mathcal{H}} = 
\bigl(
	\mathbf{d}^\top \mathbf{d} + 
	\sum_{k=1}^K (\mathbf{d}^{(k)})^\top  \boldsymbol{\Phi}^{(K)}  \mathbf{d}^{(k)} 
	\bigr)^{1/2}.
\end{equation*}
Therefore, the estimated PLS direction is the tuple $\hat{\boldsymbol{\xi}} \in \mathcal{H}$, defined as
\begin{equation}\label{def:computed_PLS_direction}
	\hat{\boldsymbol{\xi}} :=
	\frac{1}{
	\bigl(
	\mathbf{d}^\top \mathbf{d} + 
	\sum_{k=1}^K (\mathbf{d}^{(k)})^\top  \boldsymbol{\Phi}^{(K)}  \mathbf{d}^{(k)} 
	\bigr)^{1/2}
	}
	\biggl(
	\sum_{m=1}^M d^{(1)}_m \phi^{(1)}_m,
	\ldots,
	\sum_{m=1}^M d^{(K)}_m \phi^{(K)}_m,
	\mathbf{d}
	\biggr).
\end{equation}
Finally, the estimated $l$th PLS scores for the $n$ observations,
denoted $\hat{\boldsymbol{\rho}} \in \mathbb{R}^n$
 is computed simultaneously as the hybrid inner product (Definition \ref{def:n_one_inner_product}):
	\begin{equation*}
	\langle \widetilde{\mathbf{W}}, \hat{\boldsymbol{\xi}} \rangle_{\mathcal{H}}
	=
	\frac{1}{\|	\hat{\boldsymbol{\xi}} \|_{\mathcal{H}}}
	\biggl(
	\mathbf{Z} \mathbf{d}
	+
	\sum_{k=1}^K \widetilde{\mathbf{C}}^{(k)} \boldsymbol{\Phi}^{(k)} \mathbf{d}^{(k)}
	\biggr)
	. 
\end{equation*}

%	consider the basis function expansions of the functional predictors $X_i^{(k)}(t_k) = \sum_{m=1}^M c_{im}^{(k)} b^{(k)}_{m}(t_k)$, $i=1, \ldots, n$, and those of the functional parts of the PLS component $\psi^{(k)}(t_k) = \sum_{m=1}^M a^{(k)}_{m} b^{(k)}_m(t_k)$, $k=1, \ldots, K$. Let 
	\begin{proposition} \label{Proposition: generalized rayleigh quotient}
		  For any unit-length direction in $\mathcal{H}$
		  whose functional parts lie in $\operatorname{span}(\Phi^{(1)}), \ldots, \operatorname{span}(\Phi^{(K)})$,
		let us define
		the empirical and basis-approximated sample covariance between the resulting component and response as
\begin{equation*}
	\widehat{\textnormal{Cov}} (\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y )
	:=
	\frac{1}{n} 
	\mathbf{y}^\top 
 \langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_\mathcal{H},
\end{equation*}
where right-hand side notation comes from Definition \ref{def:n_one_inner_product}.
Then the PLS direction 
$	\hat{\boldsymbol{\xi}}$ 
 computed in \eqref{def:computed_PLS_direction} maximizes $\widehat{\textnormal{Cov}} (\langle \widetilde{\mathbf{W}}, \boldsymbol{\xi} \rangle_{\mathcal{H}}, Y )$.
	\end{proposition}
